---
title: "Building OpenAlex API Queries with `pro_query()`"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    mermaid:
      theme: neutral
vignette: >
  %\VignetteIndexEntry{Building OpenAlex API Queries with `pro_query()`}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  cache: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `pro_query()` function is the foundation of the `openalexPro` workflow. It constructs well-formed URLs for querying the OpenAlex API, handling parameter validation, filter construction, and automatic request chunking for large queries.

This vignette provides a comprehensive guide to using `pro_query()`, including:

- Basic usage patterns
- Advanced filtering and selection
- Automatic chunking for large queries
- Error handling and validation
- Internal architecture and flow

# Basic Usage

## Simple Query

The most basic query requires only an entity type:

```{r}
library(openalexPro)

# Query for works
url <- pro_query(entity = "works")
url
```

## Full-Text Search

Add a search term to perform full-text search:

```{r}
url <- pro_query(
  entity = "works",
  search = "climate change biodiversity"
)
url
```

## Field Selection

Use the `select` parameter to specify which fields to return:

```{r}
url <- pro_query(
  entity = "works",
  search = "machine learning",
  select = c("ids", "title", "publication_year", "cited_by_count")
)
url
```

**Note:** The function validates field names against `opt_select_fields()` and provides helpful suggestions for typos:

```{r}
# This will error with a suggestion
try(
  pro_query(
    entity = "works",
    select = c("id", "title") # Should be "ids" not "id"
  )
)
# Error: Invalid select field(s): id.
# Did you mean: id → ids?
# Valid select field(s) are defined in `opt_select_fields()`.
```

# Filtering

## Basic Filters

Filters are passed as named arguments via `...`:

```{r}
url <- pro_query(
  entity = "works",
  from_publication_date = "2020-01-01",
  to_publication_date = "2023-12-31",
  type = "article"
)
url
```

## Multiple Values (OR Logic)

Pass a vector to express OR logic within a filter:

```{r}
url <- pro_query(
  entity = "works",
  language = c("en", "de", "fr"), # English OR German OR French
  type = c("article", "preprint") # Article OR Preprint
)
url
```

## Filter Validation

Filter names are validated against `opt_filter_names()`:

```{r}
# This will error with suggestions
try(
  pro_query(
    entity = "works",
    publiction_year = 2023 # Typo: should be "publication_year"
  )
)
# Error: Invalid filter name(s): publiction_year.
# Did you mean: publiction_year → publication_year?
# Valid filter names are defined in `opt_filter_names()`.
```

# Advanced Features

## Grouping (Facets)

Use `group_by` to get aggregate counts:

```{r}
url <- pro_query(
  entity = "works",
  search = "artificial intelligence",
  group_by = "publication_year"
)
url
```

## Additional Options

The `options` parameter accepts additional query parameters:

```{r}
url <- pro_query(
  entity = "works",
  search = "quantum computing",
  options = list(
    per_page = 200,
    sort = "cited_by_count:desc",
    cursor = "*"
  )
)
url
```

## Single Entity Retrieval

Fetch a specific entity by ID:

```{r}
url <- pro_query(
  entity = "works",
  id = "W2741809807"
)
url
```

# Automatic Chunking

When querying with large lists of DOIs or IDs, `pro_query()` automatically splits the request into chunks to avoid API limits.

## Chunking Behavior

By default, requests are chunked at 50 items. The following filters trigger automatic chunking:

- `openalex`
- `ids.openalex`
- `doi`
- `cites`
- `cited_by`

## Example: Large DOI List

```{r}
# Create a list of 120 DOIs
dois <- paste0("10.1234/example", 1:120)

# This returns a named list of URLs
urls <- pro_query(
  entity = "works",
  doi = dois,
  chunk_limit = 50 # Default
)

# Check the result
length(urls) # 3 chunks (50 + 50 + 20)
names(urls) # "chunk_1", "chunk_2", "chunk_3"
```

## Custom Chunk Size

Adjust the chunk size with `chunk_limit`:

```{r}
urls <- pro_query(
  entity = "works",
  doi = dois,
  chunk_limit = 100 # Larger chunks
)

length(urls) # 2 chunks (100 + 20)
```

# Supported Entities

`pro_query()` supports all major OpenAlex entities:

```{r}
# Works
pro_query(entity = "works", search = "neuroscience")

# Authors
pro_query(entity = "authors", search = "Einstein")

# Institutions
pro_query(entity = "institutions", search = "MIT")

# Venues (journals, conferences)
pro_query(entity = "venues", search = "Nature")

# Concepts (topics/fields)
pro_query(entity = "concepts", search = "machine learning")

# Publishers
pro_query(entity = "publishers", search = "Springer")

# Funders
pro_query(entity = "funders", search = "NSF")
```

# Function Architecture

## High-Level Flow

```{mermaid}
flowchart TD
    Start([pro_query called]) --> ValidateEntity[Validate entity type]
    ValidateEntity --> GatherFilters[Gather filters from ...]
    GatherFilters --> CheckMultipleID{Multiple IDs<br/>provided?}

    CheckMultipleID -->|Yes| MoveToFilter[Move IDs to<br/>ids.openalex filter]
    CheckMultipleID -->|No| ValidateInputs[Validate filters<br/>and select fields]
    MoveToFilter --> ValidateInputs

    ValidateInputs --> CheckChunking{Filters contain<br/>chunkable fields?}

    CheckChunking -->|No| SingleBatch[Single batch:<br/>filter_batches = list]
    CheckChunking -->|Yes| ChunkLoop[Split large filters<br/>into chunks]

    ChunkLoop --> CheckSize{Values > chunk_limit?}
    CheckSize -->|Yes| SplitValues[Split values into<br/>groups of chunk_limit]
    CheckSize -->|No| KeepIntact[Keep filter intact]

    SplitValues --> NextFilter{More chunkable<br/>filters?}
    KeepIntact --> NextFilter
    NextFilter -->|Yes| ChunkLoop
    NextFilter -->|No| BuildURLs

    SingleBatch --> BuildURLs[Build URLs for each batch]

    BuildURLs --> ForEachBatch[For each batch:<br/>1. Build filter string<br/>2. Build select string<br/>3. Assemble query params<br/>4. Create URL]

    ForEachBatch --> CheckCount{Multiple batches?}

    CheckCount -->|Yes| ReturnList[Return named list:<br/>chunk_1, chunk_2, ...]
    CheckCount -->|No| ReturnSingle[Return single URL]

    ReturnList --> End([Return URLs])
    ReturnSingle --> End

    style Start fill:#e1f5e1
    style End fill:#e1f5e1
    style ValidateInputs fill:#fff3cd
    style ChunkLoop fill:#cce5ff
    style BuildURLs fill:#f8d7da
```

## Internal Helper Functions

The function uses several internal helpers for clean, maintainable code:

### `.is_empty(x)`
Checks if an object is NULL or has zero length.

```{r}
# Used throughout for consistent NULL/empty checking
.is_empty(NULL) # TRUE
.is_empty(character()) # TRUE
.is_empty("value") # FALSE
```

### `.oa_collapse(x)`
Collapses vectors into pipe-separated strings for the API.

```{r}
# Handles NULL, NA values, and logical conversion
.oa_collapse(c("en", "de", "fr")) # "en|de|fr"
.oa_collapse(c(TRUE, FALSE)) # "true|false"
.oa_collapse(c("a", NA, "b")) # "a|b"
```

### `.oa_build_filter(fl)`
Constructs the filter query string from a named list.

```{r}
# Combines multiple filters with commas
filters <- list(
  from_publication_date = "2020-01-01",
  language = c("en", "de"),
  type = "article"
)
.oa_build_filter(filters)
# "from_publication_date:2020-01-01,language:en|de,type:article"
```

### `.fuzzy_suggest(bad, allowed)`
Provides spelling suggestions for invalid field names using edit distance.

```{r}
# Suggests corrections within edit distance of 3
.fuzzy_suggest("titel", c("title", "author", "year")) # "title"
.fuzzy_suggest("publiction_year", opt_filter_names()) # "publication_year"
```

### `.build_validation_error(bad, allowed, field_type, helper_fn_name)`
Constructs helpful error messages with suggestions.

### `.validate_select(select)` and `.validate_filter(fl)`
Validate field names and filter names against allowed values.

## Chunking Algorithm

The chunking algorithm is designed to handle multiple chunkable filters simultaneously:

```{mermaid}
flowchart TD
    Start([Start with filter_batches<br/>containing original filter]) --> IdentifyTargets[Identify chunk targets<br/>in current filters]

    IdentifyTargets --> LoopTargets{For each<br/>chunk target}

    LoopTargets --> LoopBatches{For each<br/>current batch}

    LoopBatches --> GetValues[Get values for<br/>this target key]
    GetValues --> RemoveNA[Remove NA values]

    RemoveNA --> CheckSize{length > chunk_limit?}

    CheckSize -->|Yes| Split[Split into chunks<br/>of chunk_limit]
    CheckSize -->|No| Keep[Keep batch as-is]

    Split --> CreateBatches[Create new batches<br/>with chunked values]
    Keep --> AddToNew[Add to new_batches]
    CreateBatches --> AddToNew

    AddToNew --> MoreBatches{More batches<br/>to process?}
    MoreBatches -->|Yes| LoopBatches
    MoreBatches -->|No| UpdateBatches[filter_batches = new_batches]

    UpdateBatches --> MoreTargets{More targets<br/>to process?}
    MoreTargets -->|Yes| LoopTargets
    MoreTargets -->|No| Return([Return filter_batches])

    style Start fill:#e1f5e1
    style Return fill:#e1f5e1
    style CheckSize fill:#fff3cd
    style Split fill:#f8d7da
```

### Example: Chunking Multiple Filters

```{r}
# Both DOI and cites will be chunked
urls <- pro_query(
  entity = "works",
  doi = paste0("10.1234/example", 1:75), # 75 DOIs
  cites = paste0("W", 1000:1089) # 90 IDs
)

# Result: 75/50 = 2 DOI chunks, 90/50 = 2 cites chunks
# Cross product: 2 × 2 = 4 total URL batches
length(urls) # 4
```

# Best Practices

## 1. Use Field Selection

Always use `select` to limit returned fields and reduce API response size:

```{r}
# Good: Only request needed fields
pro_query(
  entity = "works",
  select = c("ids", "title", "publication_year")
)

# Avoid: Requesting all fields (default behavior without select)
```

## 2. Leverage Chunking

Let `pro_query()` handle large ID lists automatically:

```{r}
# The function handles chunking transparently
urls <- pro_query(
  entity = "works",
  doi = large_doi_vector # Even thousands of DOIs
)

# Then iterate over chunks in your request logic
for (url in urls) {
  # Process each chunk...
}
```

## 3. Validate Early

Use the validation helpers to check your filters before making requests:

```{r}
# Check available filters
opt_filter_names()

# Check available select fields
opt_select_fields()
```

## 4. Use Sorting and Pagination

For large result sets, use appropriate options:

```{r}
pro_query(
  entity = "works",
  search = "climate change",
  options = list(
    per_page = 200, # Maximum page size
    sort = "publication_date:desc",
    cursor = "*" # For cursor pagination
  )
)
```

# Common Use Cases

## Finding Recent Publications

```{r}
url <- pro_query(
  entity = "works",
  from_publication_date = "2024-01-01",
  type = "article",
  select = c("ids", "title", "publication_date", "authorships"),
  options = list(
    per_page = 200,
    sort = "publication_date:desc"
  )
)
```

## Analyzing Highly Cited Works

```{r}
url <- pro_query(
  entity = "works",
  from_cited_by_count = 100,
  type = "article",
  select = c("ids", "title", "cited_by_count", "publication_year"),
  options = list(
    sort = "cited_by_count:desc",
    per_page = 50
  )
)
```

## Finding Author Publications

```{r}
url <- pro_query(
  entity = "works",
  `author.id` = "A2208157607", # Use backticks for dots in names
  select = c("ids", "title", "publication_year"),
  options = list(sort = "publication_date:desc")
)
```

## Institution Output Analysis

```{r}
url <- pro_query(
  entity = "works",
  `institutions.id` = "I4200000001",
  from_publication_date = "2020-01-01",
  type = "article",
  select = c("ids", "title", "authorships", "publication_year")
)
```

## Bulk DOI Lookup

```{r}
# Read DOIs from file
dois <- readLines("my_dois.txt")

# Query in chunks automatically
urls <- pro_query(
  entity = "works",
  doi = dois,
  select = c("ids", "title", "cited_by_count")
)

# Process each chunk
results <- lapply(urls, function(url) {
  # Use pro_request() or similar to fetch data
  # pro_request(url)
})
```

# Error Handling

The function provides helpful error messages for common mistakes:

## Invalid Entity

```{r}
try(pro_query(entity = "paper"))
# Error: 'arg' should be one of "works", "authors", "venues",
#        "institutions", "concepts", "publishers", "funders"
```

## Invalid Filter Names

```{r}
try(pro_query(entity = "works", invalid_filter = "value"))
# Error: Invalid filter name(s): invalid_filter.
# Valid filter names are defined in `opt_filter_names()`.
```

## Invalid Select Fields

```{r}
try(pro_query(entity = "works", select = c("id", "titel")))
# Error: Invalid select field(s): id, titel.
# Did you mean: id → ids, titel → title?
# Valid select fields are defined in `opt_select_fields()`.
```

# Integration with openalexPro Workflow

`pro_query()` is the first step in the typical `openalexPro` workflow:

```{r}
library(openalexPro)

# Step 1: Build query
urls <- pro_query(
  entity = "works",
  search = "machine learning healthcare",
  from_publication_date = "2020-01-01",
  type = "article",
  select = c("ids", "title", "abstract", "publication_year")
)

# Step 2: Retrieve data
# pro_request(urls, output_dir = "data/json")

# Step 3: Convert to JSONL
# pro_request_jsonl("data/json", "data/jsonl")

# Step 4: Convert to Parquet
# pro_request_jsonl_parquet("data/jsonl", "data/parquet")

# Step 5: Query with DuckDB
# ... analyze your data ...
```

# See Also

- `opt_filter_names()` - List available filter names
- `opt_select_fields()` - List available select fields
- `pro_request()` - Execute API requests and save results
- `pro_request_jsonl()` - Convert JSON to JSONL format
- `pro_request_jsonl_parquet()` - Convert JSONL to Parquet format

# References

- [OpenAlex API Documentation](https://docs.openalex.org/)
- [OpenAlex Filters](https://docs.openalex.org/how-to-use-the-api/get-lists-of-entities/filter-entity-lists)
- [OpenAlex Select Fields](https://docs.openalex.org/how-to-use-the-api/get-lists-of-entities/select-fields)
