---
title: "Working with OpenAlex Snapshots"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Working with OpenAlex Snapshots}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  cache: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

OpenAlex provides complete data snapshots that can be downloaded and processed locally. This approach offers several advantages over API queries:
- **No rate limits**: Process millions of records without API restrictions
- **Reproducibility**: Work with a fixed point-in-time dataset
- **Speed**: Local queries are significantly faster than API calls (in many cases, especially what larger result corp[ora concerns)])
- **Offline access**: No internet connection required after download

This vignette guides you through:

1. Setting up your environment
2. Downloading the OpenAlex snapshot
3. Converting to parquet format
4. Building indexes for fast lookups
5. Querying your local corpus

# Prerequisites

## Hardware Requirements

Working with the full OpenAlex snapshot requires significant resources:

| Resource | Minimum | Recommended |
|----------|---------|-------------|
| Disk space | 2.5 TB | 3+ TB |
| RAM | 16 GB | 32+ GB |
| CPU | 2 cores | 4+ cores |

The snapshot download is approximately 1.2 TB. The parquet conversion adds another 100-200 GB depending on compression settings.

## Software Requirements

- **R** (>= 4.1.2)
- **AWS CLI**: For downloading from S3
- **GNU Make**: For running the Makefile targets
- **openalexPro**: This package

### Installing AWS CLI

```bash
# macOS (with Homebrew)
brew install awscli

# Ubuntu/Debian
sudo apt install awscli

# Windows (with winget)
winget install Amazon.AWSCLI
```

Verify installation:

```bash
aws --version
```

No AWS account is required - the OpenAlex bucket allows anonymous access.

# Quick Start

The easiest way to get started is using `prepare_snapshot()`:
```{r}
library(openalexPro)

# Prepare a directory for your snapshot
prepare_snapshot("/path/to/openalex-data")
```

This creates:

- `Makefile`: Contains all commands for snapshot management
- `snapshot_guide.html`: This documentation

Then navigate to the directory and use Make:

```bash
cd /path/to/openalex-data

# See available commands
make help

# Download snapshot (WARNING: ~350 GB, takes hours)
make snapshot

# Convert to parquet format
make parquet

# Build search indexes
make parquet_index
```

# Step-by-Step Guide

## 1. Prepare Your Directory

Choose a location with sufficient disk space:

```{r}
library(openalexPro)

# Create and prepare the snapshot directory
snapshot_dir <- "/Volumes/external-drive/openalex"
prepare_snapshot(snapshot_dir)
```

## 2. Download the Snapshot

The OpenAlex snapshot is hosted on Amazon S3 and updated regularly. The download uses `aws s3 sync` which:

- Downloads only new or changed files
- Resumes interrupted downloads
- Removes files deleted from the source

```bash
cd /Volumes/external-drive/openalex
make snapshot
```

**Expected time**: 2-12 hours depending on your internet connection.

**Tip**: You can monitor progress in another terminal:

```bash
watch -n 60 'du -sh openalex-snapshot/*'
```

### Snapshot Structure

After download, you'll have:

```
openalex-snapshot/
├── RELEASE_NOTES.txt
├── data/
│   ├── works/
│   ├── authors/
│   ├── institutions/
│   ├── sources/
│   ├── publishers/
│   ├── funders/
│   ├── topics/
│   ├── fields/
│   ├── subfields/
│   └── domains/
└── ...
```

Each entity type contains gzipped JSON files partitioned by update date.

## 3. Convert to Parquet Format

The JSON snapshot is not efficient for analytical queries. Converting to parquet format provides:

- **Columnar storage**: Read only the columns you need
- **Compression**: Smaller files on disk
- **Fast filtering**: Predicate pushdown for efficient queries
- **Type safety**: Proper data types for each field

```bash
make parquet
```

This runs `snapshot_to_parquet()` which:

1. Reads all JSON files for each entity type
2. Flattens nested structures where appropriate
3. Writes partitioned parquet files

**Expected time**: 2-8 hours depending on CPU and disk speed.

### Parquet Structure

After conversion:

```
parquet/
├── works/
│   ├── part-00000.parquet
│   ├── part-00001.parquet
│   └── ...
├── authors/
├── institutions/
└── ...
```

## 4. Build Search Indexes

For fast ID-based lookups, build indexes:

```bash
make parquet_index
```

This creates Hive-partitioned indexes that allow O(1) lookups by OpenAlex ID:

```
parquet/
├── works/
├── works_id_idx/
│   ├── id_block=0/
│   │   └── data.parquet
│   ├── id_block=1/
│   │   └── data.parquet
│   └── ...
└── ...
```

The index partitions records by ID blocks (groups of 10,000 IDs), enabling direct access to specific records without scanning the entire corpus.

# Using Your Local Corpus

## Looking Up Records by ID

```{r}
library(openalexPro)

# Look up specific works by OpenAlex ID
works <- lookup_by_id(
  index_dir = "/Volumes/external-drive/openalex/parquet/works_id_idx",
  ids = c("W2741809807", "W2100837269")
)
```

## Looking Up by DOI

First, build a DOI index:

```{r}
build_corpus_index(
  corpus_dir = "/Volumes/external-drive/openalex/parquet/works",
  index_dir = "/Volumes/external-drive/openalex/parquet/works_doi_idx",
  id_column = "doi"
)
```

Then query:

```{r}
works <- lookup_by_id(
  index_dir = "/Volumes/external-drive/openalex/parquet/works_doi_idx",
  ids = "https://doi.org/10.1038/nature12373",
  id_column = "doi"
)
```

## Direct DuckDB Queries

For complex analytical queries, use DuckDB directly:

```{r}
library(duckdb)
library(DBI)

con <- dbConnect(duckdb())

# Register parquet files as a view
dbExecute(
  con,
  "
  CREATE VIEW works AS
  SELECT * FROM parquet_scan('/Volumes/external-drive/openalex/parquet/works/*.parquet')
"
)

# Run analytical queries
result <- dbGetQuery(
  con,
  "
  SELECT publication_year, COUNT(*) as n_works
  FROM works
  WHERE publication_year >= 2020
  GROUP BY publication_year
  ORDER BY publication_year
"
)

dbDisconnect(con)
```

# Managing Updates

OpenAlex releases new snapshots regularly. The Makefile handles updates gracefully:

1. **Existing data is preserved**: Before downloading a new snapshot, the old one is renamed with its release date (e.g., `openalex-snapshot-2026-01-15`)

2. **Incremental sync**: `aws s3 sync` only downloads changed files

3. **Re-conversion**: After downloading, run `make parquet` again to update the parquet files

## Checking for Updates

```bash
make snapshot_info
```

This shows the current state of the S3 bucket including the total size and file count.

## Manual Update Workflow

```bash
# Check current release date
head -5 openalex-snapshot/RELEASE_NOTES.txt

# Sync new data (old snapshot is automatically renamed)
make snapshot

# Check new release date
head -5 openalex-snapshot/RELEASE_NOTES.txt

# Re-convert to parquet (old parquet directory is automatically renamed)
make parquet

# Rebuild indexes
make parquet_index
```

# Troubleshooting

## Download Issues

**Interrupted download**: Simply run `make snapshot` again. The sync will resume where it left off.

**Slow download**: OpenAlex uses CloudFront CDN. Speed depends on your location and internet connection. Consider running overnight.

**Disk space errors**: Ensure you have at least 400 GB free before starting.

## Conversion Issues

**Out of memory**: Reduce memory usage in the Makefile:

```makefile
parquet:
	Rscript \
	-e 'openalexPro::snapshot_to_parquet(..., memory_limit = "4GB", threads = 1)'
```

**DuckDB errors**: Ensure you have the latest version of the `duckdb` R package.

## Index Issues

**Slow lookups**: Ensure `merge_partitions = TRUE` was used when building the index. This creates single files per partition for faster reads.

# Makefile Reference

| Target | Description |
|--------|-------------|
| `help` | Show available targets |
| `snapshot_info` | Display S3 bucket information |
| `snapshot` | Download/sync snapshot from S3 |
| `parquet` | Convert snapshot to parquet |
| `parquet_index` | Build ID indexes |
| `clean_parquet` | Remove parquet directory |
| `clean_snapshot` | Remove snapshot directory |
| `clean` | Remove both directories |

## Customizing the Makefile

Edit the variables at the top of the Makefile:

```makefile
SNAPSHOTDIR=./openalex-snapshot   # Where to download JSON
PARQUETDIR=./parquet                   # Where to write parquet
```

You can also adjust memory and thread settings in the `parquet` and `parquet_index` targets.

# Additional Resources

- [OpenAlex Documentation](https://docs.openalex.org/)
- [OpenAlex Snapshot Documentation](https://docs.openalex.org/download-all-data/openalex-snapshot)
- [DuckDB Documentation](https://duckdb.org/docs/)
- [Apache Parquet Format](https://parquet.apache.org/)
