---
title: "Workflow using `openalexPro`"
format:
  pdf:
    toc: false
vignette: >
  %\VignetteIndexEntry{Vignette's Title}
  %\VignetteEngine{quarto::pdf}
  %\VignetteEncoding{UTF-8}
execute: 
  cache: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The package `openalexPro` is an extension of the package `openalexR`. The main difference is that `openalexR` does all processing in memory and returns the result as an R object, while `openalexPro` will write the results to disk. This allows for processing very large corpora.

Most of the processing is done using the `duckdb` package which makes it possible tro do complex queries of the corpora without the need to load the whole corpus into memory.

# Workflow

The workflow when using `openalexPro` is essentially as follows:

1. the function `openalexR::oa_query()` is used to build the API query for the OpenAlex API
2. the function `openalexPro::pro_request()` is used to retrieve the results from the OpenAlex API and to store them in a folder in the format oif `json` files as returned by OpenAlex.
3. the function `openalexPro::pro_request_json_to_parquet()` is used to convert the `json` files to a parquet database


TODO: Add plantuml graph



```{r setup}
library(openalexPro)
```
